Toward \earlywarning\ detection of gravitational waves from compact binary coalescence

### 2011-08-14 article.tex
Put back in real time claims, re-flowed some TeX code
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1347 ccdd05d5-56c1-461a-bf98-85f6876523b4

\section{Introduction}
\label{sec:introduction}

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  If a
neutron star (\NS{}) is involved, it might become tidally disrupted near
the merger and fuel an electromagnetic (\EM{}) counterpart
\citep{shibata:2007}.  Effort from both the \GW{} and the broader astronomical communities might make it
possible to use \GW{} observations as an early warning trigger for \EM{}
followup. In the first generation of ground-based laser interferometers, the
\GW{} community initiated a project to send alerts when potential \GW{}
transients were observed in order to trigger followup observations by \EM{}
telescopes.  The typical latencies were 30 minutes \citep{HugheyGWPAW2011},
which was an important achievement, but too late to catch any prompt \EM{}
emission.  Since the \GW\ signal is in principle detectable even before the tidal
disruption, one might have the ambition of reporting \GW\ candidates not minutes
after the merger, but seconds before.  We explore one essential ingredient of this
problem, a computationally inexpensive low-latency filtering algorithm for detecting
inspiral signals in \GW{} data.  We also consider the prospects for advanced \GW{}
detectors and discuss other areas of work that would be required for rapid analysis.

Compact binary coalescence (\CBC{}) is a plausible progenitor for most short
gamma-ray bursts (short \GRB{}s) \citep{Lee:2005, nakar07}, but the
association is not iron-clad \citep{2011ApJ...727..109V}. The tidally
disrupted material falls onto the newly formed, rapidly spinning compact object
and is accelerated in jets along the spin axis with a timescale of
$0.1$--$1$~s after the merger \citep{Janka1999}, matching the short \GRB{}
duration distribution well. Prompt \EM{} emission including the \GRB{}
can arise as fast outflowing
matter collides with slower matter ejected earlier in inner shocks. The same
inner shocks, or potentially reverse shocks, can produce an accompanying
optical flash \citep{Sari99}. The prompt emission is a probe into the extreme
initial conditions of the outflow, in contrast with afterglows, which arise in
the external shock with the local medium and are relatively insensitive to
initial conditions. Optical flashes have been observed for a handful of long
\GRB{}s \citep{2011CRPhy..12..255A} by telescopes with extremely rapid response
or, in the case of GRB 080319b, by pure serendipity, where several
telescopes were already observing the afterglow of another \GRB{} in the same
field of view \citep{2008Natur.455..183R}. The observed optical flashes peaked
within tens of seconds and decayed quickly. For short \GRB{} energy balance and
plasma density,
however, the reverse shock model predicts a peak flux in radio, approximately
20 minutes after the \GRB{}, but also a relatively faint optical flash
\citep{nakar07}; for a once-per-year Advanced LIGO event at $130$~Mpc, the
radio flux will peak around 9~GHz at $\sim$$5$~mJy, with emission in the
$R$-band at $\sim$19~mag. Interestingly, roughly a quarter to half of observed short
\GRB{}s also exhibit extended X-ray emission of $30$--$100$~s in duration
beginning $\sim$$10$~s after the \GRB{} and carrying comparable fluence to the
initial outburst. This can be explained if the merger results in the formation
of a proto-magnetar that interacts with ejecta \citep{Bucciantini2011}. Rapid
\GW{} alerts would enable joint \EM{} and \GW{} observations to confirm
the short \GRB{}-\CBC{} link and allow the early \EM{}
observation of exceptionally nearby and thus bright events.

@@

The paper is organized as follows.  First, we discuss prospects for early-warning
detection.  Then, we provide an overview of our novel method for detecting compact
binary coalescence signals with extremely low latency. We then describe a prototype
implementation using open source signal processing software.  To validate our approach
we present results of simulations.  We conclude with some remarks on what remains to
prepare for the advanced detector era.

@@

\section{Novel low latency algorithm for \CBC\ detection}
\label{sec:method}

@@

Here we describe a method for reducing the computational cost of a \TD\ search
for compact binary coalescence.  We give an extremely low latency algorithm
that competes in terms of floating point operations per second with the
conventional \FD\ method, which by contrast requires a significant latency due to the
inherent acausality of the Fourier transform.  Our method, called \lloid{}
(Low Latency Online Inspiral Detection),
involves two transformations of the templates that produce a
network of orthogonal filters that is far more computationally
efficient than the original bank of matched filters.

@@

Our goal is to reduce the filtering cost of a
large fraction of the waveform by computing part of the convolution at a lower
sample rate.  Specifically we consider here time slice boundaries with the
smallest power-of-two sample rates that sub-critically sample the time-sliced
templates.  The time slices consist of the $S$ intervals
$\left[t^0, t^1\right),\, \left[t^1, t^2\right),\, \dots,\, \left[t^{S-1}, t^S\right)$,
sampled at frequencies $f^0,\, f^1,\, \dots,\, f^\mathrm{S-1}$ where $f^s$ is at
least twice the highest nonzero frequency component of any filter in the bank for the
$s$th time slice.

@@

First, the sample rate of the detector data must be decimated to match sample rates
with each of the time slices.  We will denote the decimated detector data streams using
a superscript $^s$ to indicate the time slices to which they correspond, and the
operator $H^\shortdownarrow$ will represent the appropriate decimation filter:
\begin{equation*}
\label{eq:decomp}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k].
\end{equation*}
We shall use the symbol $H^\shortuparrow$ to represent an interpolation filter,
\begin{equation*}
	x^{s}[k] = \left( H^\shortuparrow x^{s+1}\right)[k].
\end{equation*}
Ultimately, the latency of the entire \lloid\ algorithm is set by the decimation
and interpolation filters because they are generally time-symmetric and slightly
acausal.

@@

Taking into account the decimation of the detector data, the orthogonal \fir\ filters, the reconstruction of the time-sliced templates, the interpolation of \SNR\ from previous time slices, and the accumulation of \SNR, in total the \lloid\ algorithm requires
\begin{multline*}
\sum_{\mathclap{s=0}}^{\mathclap{S-1}} \left( 2 \slicessamps \numsvdtmps + 2 \numtmps \numsvdtmps + \numtmps \right) f^s \\ + 2\sum_{\mathclap{f^s \in \{f^k \, : \, 0 < k < S\}}} \left( N^\shortdownarrow f^0 + \numtmps N^\shortuparrow f^{s-1}\right)
\end{multline*}
\flops.  The second sum is carried out over the set of distinct sample rates (except for the base sample rate) rather than over the time slices themselves, as we have found that it is sometimes desirable to place multiple adjacent time slices at the same sample rate in order to keep the size of matrices that enter the singular value decomposition manageable.  Here we have assumed that the decimation filters are connected in parallel, converting from the base sample rate $f^0$ to each of the time slice sample rates $f^1$, $f^2$, $\dots$, and that the interpolation filters are connected in cascade fashion with each interpolation filter stepping from the sample rate of one time slice to the next.

@@

The latency of the \lloid\ method is entirely due to the acausality of the interpolation and decimation filters.  The latency is exactly
$$
\frac{1}{2} \left( N^\shortdownarrow / f^{S-1} + \sum_{\mathclap{f^s \in \{f^k \, : \, 0 < k < S\}}} N^\shortuparrow / f^s \right) \textrm{ seconds.}
$$
The first term is due to the decimation filter that converts from the base sample rate $f^0$ to the lowest sample rate $f^{S-1}$ and the sum that comprises the second term is due to the latency of all of the interpolation filters connected in series.

@@

\section{Implementation}
\label{sec:implementation}

@@

The \lloid\ algorithm is amenable to extremely low latency implementation.
The lowest possibly latency could be achieved by integration directly into
the data acquisition and storage system of the \LIGO\ observatories.  A slightly more 
modest goal is to leverage existing signal processing software in order to implement
the \lloid\ algorithm.

@@

The whitened detector data is reduced to successively
lower sample rates by decimation. Decimation involves applying an anti-aliasing
filter to the data, and then down-sampling by deleting samples.  We use GStreamer's
stock {\tt audioresample} element, which provides a FIR decimation filter with
a Kaiser-windowed sinc function kernel.

@@

Next, these decimated data streams are fed into parallel \fir\ filter banks.
The basis filters are implemented using a \gstlal\ element called {\tt
lal\_firbank}, which provides a single input, multiple output bank of \fir\ filters
from a matrix of filter coefficients.  Several instances of {\tt lal\_firbank}
execute in parallel to handle each of the time slices.  Filter coefficients are taken
from the nonzero samples of the time slices in equation~\eqref{eq:time-sliced-templates},
with the zero-padding effectively accomplished by appropriate queues, or delay lines.

@@

In order to form the early-warning output from each time slice, we have to add
the partial \SNR\ to the early-warning output from the subsequent time slices.
If the next time slice does not have the same sample rate, its output must
first be interpolated.  The \gstreamer\ element {\tt audioresample} is used again here.

@@

\section{Results}
\label{sec:results}

@@

We examine the performance of the \lloid\ algorithm on a small region of
compact binary parameter space centered on typical \NS--\NS{}
masses.  We begin by constructing a template bank that spans component masses
from 1~to~3~$M_\odot$ using a simulated Advanced \LIGO{} noise
curve~\citep{ALIGONoise}.  Templates are truncated at 10~Hz, where the projected
sensitivity of Advanced \LIGO\ is interrupted by the ``seismic wall.''
This results in a grid of 98544 points, or
$2 \times 98544 = 197088$~templates.  Then we create sub-banks by partitioning
the parameter space by chirp mass.  We shall concentrate on a sub-bank with 657 points with
chirp masses between 1.1955 and 1.2045~$M_\odot$, or $2 \times 657 = 1314$~templates.
Figure \ref{fig:tmpltbank} illustrates this procedure.
\begin{figure}[h]
	\plotone{f4}
	\caption{\label{fig:tmpltbank}Source parameters selected for sub-bank used in this
case study, consisting of component masses $m_1$, $m_2$, between 1 and 3~$M_\odot$, and
chirp masses $\mathcal{M}$ between 1.1955 and 1.2045~$M_\odot$.}
\end{figure}
With this sub-bank we were able to construct an efficient time-slice decomposition
that consisted of 11 time slices with sample rates between 32 and 4096 Hz summarized
in Table~\ref{tab:time_slices}.
\begin{table*}
\caption{\label{tab:time_slices} Filter design sub-bank of 1314 templates.  From left to right, this table shows the sample rate, time interval, number of samples, and number of orthogonal templates for each time slice.  We vary \SVD{} tolerance from $\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.}
\begin{center}
\begin{minipage}[c]{0.4\textwidth}
\includegraphics{t3}
\end{minipage}
\begin{minipage}[c]{0.55\textwidth}
\begin{tabular}{rr@{,\,}lc*{6}{r}}
\tableline\tableline
\\ [-2ex]
$f^s$ & $[t^s$&$t^{s+1})$ & &\multicolumn{6}{c}{$-\log_{10}$ (1$-$\SVD{} tolerance)} \\% [1ex]
\cline{5-10}
\\[-2.5ex]
(Hz) & \multicolumn{2}{c}{(s)} & $N^s$ & $1$ & $2$ & $3$ & $4$ & $5$ & $6$ \\ \tableline
4096 & [0&0.5) & 2048 & 1 & 4 & 6 & 8 & 10 & 14 \\
512 & [0.5&4.5) & 2048 & 2 & 6 & 8 & 10 & 12 & 16 \\
256 & [4.5&12.5) & 2048 & 2 & 6 & 8 & 10 & 12 & 15 \\
128 & [12.5&76.5) & 8192 & 6 & 20 & 25 & 28 & 30 & 32 \\
64 & [76.5&140.5) & 4096 & 1 & 8 & 15 & 18 & 20 & 22 \\
64 & [140.5&268.5) & 8192 & 1 & 7 & 21 & 25 & 28 & 30 \\
64 & [268.5&396.5) & 8192 & 1 & 1 & 15 & 20 & 23 & 25 \\
32 & [396.5&460.5) & 2048 & 1 & 1 & 3 & 9 & 12 & 14 \\
32 & [460.5&588.5) & 4096 & 1 & 1 & 7 & 16 & 18 & 21 \\
32 & [588.5&844.5) & 8192 & 1 & 1 & 8 & 26 & 30 & 33 \\
32 & [844.5&1100.5) & 8192 & 1 & 1 & 1 & 12 & 20 & 23 \\
\tableline
\end{tabular}
\end{minipage}
\end{center}
\end{table*}
We use this sub-bank and decomposition for the remainder of this section.

@@

During testing, we were surprised that careful design of the decimation and interpolation stages made a crucial difference in terms of computational overhead and latency.  When we tested the pipeline with the largest interpolation filters provided by GStreamer's audioresample element, we found that rather than making a negligible contribution to the total \flops\ budget and total measured overhead they became the dominant contribution.  The largest interpolation filters also incurred latencies of several seconds.  We now appreciate that the efficiency and low latency of \lloid\ hinge on proper specification of $N^\shortuparrow$.

There is possibly further room for optimization here.  We have considered designing our own custom decimation and interpolation filters or tuning these filters separately for each time slice.

\begin{figure*}[b]
	\begin{minipage}[t]{0.5\textwidth}
		\begin{center}
			\includegraphics{f5a}
			(a) Mismatch versus \SVD\ tolerance
		\end{center}
	\end{minipage}
	\begin{minipage}[t]{0.5\textwidth}
		\begin{center}
			\includegraphics{f5b}
			(b) Mismatch versus $N^\shortuparrow$
		\end{center}
	\end{minipage}
	\caption{\label{fig:bw}Box-and-whisker plot of mismatch between nominal
template bank and \lloid\ measured impulse responses.  The upper and lower boundaries of
the boxes show the upper and lower quartiles; the lines in the center denote the medians.
The whiskers represent the minimum and maximum mismatch over all templates.  In 
(a) the interpolation filter length is held fixed at $N^\shortuparrow = 192$, while
the \SVD\ tolerance is varied from $\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.
In (b), the \SVD\ tolerance is fixed at $\left(1-10^{-6}\right)$ while $N^\shortuparrow$
is varied from 8 to 192 coefficients.}
\end{figure*}

@@

We are now prepared to offer the estimated computational cost of filtering this
sub-bank of templates compared to other methods.  We used the results of the
previous subsections to set the \SVD\ tolerance to $\left(1-10^{-4}\right)$,
the interpolation filter length to 16, and the decimation filter length to 48.
Table~\ref{table:flops} shows the
computational cost in \flops\ for the sub-bank we described above.  For the \FD\ method, an \fft\
block size of $\fftblock = 2 \tmpsamps$ is assumed, resulting in a latency of
$\left(\tmpsamps / f^0\right)$ seconds.  Both the \FD\ method and \lloid\ are
five orders of magnitude faster than the conventional \TD\ method.  However,
the \FD\ method has a latency of over half of an hour, whereas \lloid\ has a
latency of just over one second.
%
\begin{table*}
\caption{\label{table:flops}Computational cost in \flops\ and latency in seconds
of the \TD\ method, the \FD\ method, and \lloid.  Cost is given
for both the sub-bank described in section~\ref{sec:bank-setup} and a full
\NS--\NS\ search.  The last column gives the approximate number of machines per
detector required for a full advanced LIGO \NS--\NS\ search.}
\begin{center}
\begin{tabular}{lllll}
\tableline\tableline
& \flops\ & & \flops\ & number of \\
method & (sub-bank) & latency (s) & (\NS--\NS) & machines \\[0.1em]
\tableline
time domain & $4.9\times10^{13}$ & 0 & $3.8\times10^{15}$ & $\sim$$10^5$ \\
frequency domain & $5.2\times10^8$ & $2\times10^3$ & $5.9\times10^{10}$ & $\sim$$10$ \\
\lloid\ (theory) & $6.6\times10^8$ & $1.2$ & $\phantom{5.7}\sim 10^{10}$ & $\sim$$10$ \\
\lloid\ (prototype) & (0.9 cores) & $1.5$ & ------------ & $\sim$$10$ \\
\tableline
\end{tabular}
\end{center}
\end{table*}

@@

We wrote a second, stripped pipeline to evaluate the actual latency and
computational overhead.  We executed this pipeline on one of the submit
machines of the \LIGO-Caltech cluster, a Sun Microsystems Sun
Fire\texttrademark\ X4600~M2 server with eight quad-core 2.7~GHz AMD
Opteron\texttrademark\ 8384 processors.  This test consumed $\sim$90\% of the
capacity of just one out of the 32 cores, maintaining a constant latency of
$\sim$1.5~s.

@@

Since in this benchmark, the pipeline's input was buffered in
0.25~s chunks, the total measured latency of 1.5~s is entirely consistent with
1.2~s of latency being attributed to the acausality of the decimation and
interpolation filters, as calculated in section~\ref{sec:lloid-method}.
Our implementation of the \lloid\ algorithm has achieved latencies that would
make other aspects of the analysis the limiting steps assuming that the human
vetting process is automated.

\section{Conclusions}
\label{sec:conclusions}

@@

The \lloid{} method is as fast as standard \fft{} convolutions but allows for
extremely low latency operation.  We anticipate requiring $\sim$40
modern multi-core computers to search for binary neutron stars using
coincident \GW\ data from a 4-detector network.  In the future, additional
computational savings could be acheived by conditionally reconstructing the
\SNR{} time-series only during times when a composite detection statistic
crosses a threshold~\citep{svd-compdetstat}.  However, the anticipated required
number of computers is well within the current computing capabilities of the
\LIGO{} Data Grid.

@@

\LIGO\ was constructed by the California Institute of Technology and
Massachusetts Institute of Technology with funding from the National Science
Foundation and operates under cooperative agreement
PHY-0107417.  CH would like to thank Ilya Mandel for
many discussions about rate estimates and the prospects of early detection,
Patrick Brady for countless fruitful conversations about
low latency analysis methods, and John Zweizig for
disscussions about \LIGO\ data acquisition.  NF would like to thank Alessandra
Corsi and Larry Price for illuminating discussions on astronomical motivations.
This research
is supported by the National Science Foundation through a Graduate Research
Fellowship to LS and by the Perimeter Institute for Theoretical Physics through
a fellowship to CH. DK is supported from the Max Planck Gesellschaft.

>>>>

\section{Introduction}

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  If a
neutron star (\NS) is involved, it might become tidally disrupted near the
merger and fuel an electromagnetic (\EM) counterpart \citep{shibata:2007}.
Effort from both the \GW{} and the broader astronomical communities might make
it possible to use \GW\ observations as an early warning trigger for \EM{}
followup. In the first generation of ground-based laser interferometers, the
\GW\ community initiated a project to send alerts when potential \GW{}
transients were observed in order to trigger followup observations by \EM{}
telescopes.  The typical latencies were 30 minutes \citep{HugheyGWPAW2011},
which was an important achievement, but too late to catch any prompt \EM{}
emission.  Since the \GW\ signal is in principle detectable even before the tidal
disruption, one might have the ambition of reporting \GW\ candidates not minutes
after the merger, but seconds before.  We explore one essential ingredient of
this problem, a computationally inexpensive low-latency filtering algorithm for
detecting inspiral signals in \GW\ data.  We also consider the prospects for
advanced \GW\ detectors and discuss other areas of work that would be required
for rapid analysis.

Compact binary coalescence (\CBC{}) is a plausible progenitor for most short
gamma-ray bursts (short \GRB{}s) \citep{Lee:2005, nakar07}, but the
association is not iron-clad \citep{2011ApJ...727..109V}. The tidally
disrupted material falls onto the newly formed, rapidly spinning compact object
and is accelerated in jets along the spin axis with a timescale of $0.1$--$1$~s
after the merger \citep{Janka1999}, matching the short \GRB\ duration
distribution well. Prompt \EM{} emission including the \GRB\ can arise as fast
outflowing matter collides with slower matter ejected earlier in inner shocks.
The same inner shocks, or potentially reverse shocks, can produce an
accompanying optical flash \citep{Sari99}. The prompt emission is a probe into
the extreme initial conditions of the outflow, in contrast with afterglows,
which arise in the external shock with the local medium and are relatively
insensitive to initial conditions. Optical flashes have been observed for a
handful of long \GRB{}s \citep{2011CRPhy..12..255A} by telescopes with extremely
rapid response or, in the case of GRB 080319b, by pure serendipity, where
several telescopes were already observing the afterglow of another \GRB{} in
the same field of view \citep{2008Natur.455..183R}. The observed optical flashes
peaked within tens of seconds and decayed quickly. For short \GRB{} energy
balance and plasma density, however, the reverse shock model predicts a peak
flux in radio, approximately 20 minutes after the \GRB{}, but also a relatively
faint optical flash \citep{nakar07}; for a once-per-year Advanced LIGO event at
$130$~Mpc, the radio flux will peak around 9~GHz at $\sim$$5$~mJy, with emission
in the $R$-band at $\sim$19~mag. Interestingly, roughly a quarter to half of
observed short \GRB{}s also exhibit extended X-ray emission of $30$--$100$~s in
duration beginning $\sim$$10$~s after the \GRB{} and carrying comparable fluence
to the initial outburst. This can be explained if the merger results in the
formation of a proto-magnetar that interacts with
ejecta~\citep{Bucciantini2011}. Rapid \GW\ alerts would enable joint \EM\ and
\GW\ observations to confirm the short \GRB{}-\CBC{} link and allow the early
\EM\ observation of exceptionally nearby and thus bright events.

@@

The paper is organized as follows.  First, we discuss prospects for early-warning
detection.  Then, we provide an overview of our novel method for detecting compact
binary coalescence signals near real-time. We then describe a prototype
implementation using open source signal processing software.  To validate our approach
we present results of simulations.  We conclude with some remarks on what remains to
prepare for the advanced detector era.

@@

\section{Novel \realtime\ algorithm for \CBC\ detection}
\label{sec:method}

@@

Here we describe a method for reducing the computational cost of a \TD\ search
for compact binary coalescence.  We give a real-time algorithm
that competes in terms of floating point operations per second with the
conventional \FD\ method, which by contrast requires a significant latency due to the
inherent acausality of the Fourier transform.  Our method, called \lloid{}
(Low Latency Online Inspiral Detection),
involves two transformations of the templates that produce a
network of orthogonal filters that is far more computationally
efficient than the original bank of matched filters.

@@

Our goal is to reduce the filtering cost of a
large fraction of the waveform by computing part of the convolution at a lower
sample rate.  Specifically we consider here time slice boundaries with the
smallest power-of-two sample rates that sub-critically sample the time-sliced
templates.  The time slices consist of the $S$ intervals
$\left[t^0, t^1\right),\, \left[t^1, t^2\right),\, \dots,\, \left[t^{S-1}, t^S\right)$,
sampled at frequencies $f^0,\, f^1,\, \dots,\, f^\mathrm{S-1}$, where $f^s$ is at
least twice the highest nonzero frequency component of any filter in the bank for the
$s$th time slice.

@@

First, the sample rate of the detector data must be decimated to match sample rates
with each of the time slices.  We will denote the decimated detector data streams using
a superscript $^s$ to indicate the time slices to which they correspond, and the
operator $H^\shortdownarrow$ will represent the appropriate decimation filter:
\begin{equation*}
\label{eq:decomp}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k].
\end{equation*}
We shall use the symbol $H^\shortuparrow$ to represent an interpolation filter,
\begin{equation*}
	x^{s}[k] = \left( H^\shortuparrow x^{s+1}\right)[k].
\end{equation*}
Ultimately, the latency of the entire \lloid\ algorithm is set by the decimation
and interpolation filters because they are generally time-symmetric and slightly
acausal.  As long as the latency introduced by the decimation and interpolation
filters for any time slice $s$ is less than that time slice's delay $t^s$, the
total latency of the \lloid\ algorithm will be zero.

@@

Taking into account the decimation of the detector data, the orthogonal \fir\
filters, the reconstruction of the time-sliced templates, the interpolation of
\SNR\ from previous time slices, and the accumulation of \SNR, in total the
\lloid\ algorithm requires
%
\begin{multline*}
\sum_{\mathclap{s=0}}^{\mathclap{S-1}} \left( 2 \slicessamps \numsvdtmps + 2 \numtmps \numsvdtmps + \numtmps \right) f^s \\ + 2\sum_{\mathclap{f^s \in \{f^k \, : \, 0 < k < S\}}} \left( N^\shortdownarrow f^0 + \numtmps N^\shortuparrow f^{s-1}\right)
\end{multline*}
%
\flops.  The second sum is carried out over the set of distinct sample rates
(except for the base sample rate) rather than over the time slices themselves,
as we have found that it is sometimes desirable to place multiple adjacent time
slices at the same sample rate in order to keep the size of matrices that enter
the singular value decomposition manageable.  Here we have assumed that the
decimation filters are connected in parallel, converting from the base sample
rate $f^0$ to each of the time slice sample rates $f^1$, $f^2$, $\dots$, and
that the interpolation filters are connected in cascade fashion with each
interpolation filter stepping from the sample rate of one time slice to the
next.

@@

@@

\section{Implementation}

@@

The \lloid\ algorithm is amenable to latency free, real-time implementation.  However, such a real-time search pipeline would require integration directly into the data acquisition and storage systems of the \LIGO\ observatories.  A slightly more 
modest goal is to leverage existing low-latency, but not real-time, signal processing software in order to implement
the \lloid\ algorithm.

@@

The whitened detector data is reduced to successively lower sample rates by
decimation. Decimation involves applying an anti-aliasing filter to the data,
and then down-sampling by deleting samples.  We use GStreamer's stock
\texttt{audioresample} element, which provides a FIR decimation filter with
a Kaiser-windowed sinc function kernel.

@@

Next, these decimated data streams are fed into parallel \fir\ filter banks.
The basis filters are implemented using a \gstlal\ element called
\texttt{lal\_firbank}, which provides a single input, multiple output bank of
\fir\ filters from a matrix of filter coefficients.  Several instances of
\texttt{lal\_firbank} execute in parallel to handle each of the time slices.
Filter coefficients are taken from the nonzero samples of the time slices in
equation~\eqref{eq:time-sliced-templates}, with the zero-padding effectively
accomplished by appropriate queues, or delay lines.

@@

In order to form the early-warning output from each time slice, we have to add
the partial \SNR\ to the early-warning output from the subsequent time slices.
If the next time slice does not have the same sample rate, its output must first
be interpolated.  The \gstreamer\ element \texttt{audioresample} is used again here.

@@

\section{Results}

@@

We examine the performance of the \lloid\ algorithm on a small region of
compact binary parameter space centered on typical \NS--\NS{}
masses.  We begin by constructing a template bank that spans component masses
from 1~to~3~$M_\odot$ using a simulated Advanced \LIGO{} noise
curve~\citep{ALIGONoise}.  Templates are truncated at 10~Hz, where the projected
sensitivity of Advanced \LIGO\ is interrupted by the ``seismic wall.''
This results in a grid of 98544 points, or
$2 \times 98544 = 197088$~templates.  Then we create sub-banks by partitioning
the parameter space by chirp mass.  We shall concentrate on a sub-bank with 657 points with
chirp masses between 1.1955 and 1.2045~$M_\odot$, or $2 \times 657 = 1314$~templates.
Figure \ref{fig:tmpltbank} illustrates this procedure.
\begin{figure}[h]
	\plotone{f4}
	\caption{\label{fig:tmpltbank}Source parameters selected for sub-bank used in this
case study, consisting of component masses $m_1$, $m_2$, between 1 and 3~$M_\odot$, and
chirp masses $\mathcal{M}$ between 1.1955 and 1.2045~$M_\odot$.}
\end{figure}
With this sub-bank we were able to construct an efficient time-slice decomposition
that consisted of 11 time slices with sample rates between 32 and 4096 Hz summarized
in Table~\ref{tab:time_slices}.
\begin{table*}
\caption{\label{tab:time_slices} Filter design sub-bank of 1314 templates.  From
left to right, this table shows the sample rate, time interval, number of
samples, and number of orthogonal templates for each time slice.  We vary \SVD\
tolerance from $\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.}
\begin{center}
\begin{minipage}[c]{0.4\textwidth}
\includegraphics{t3}
\end{minipage}
\begin{minipage}[c]{0.55\textwidth}
\begin{tabular}{rr@{,\,}lc*{6}{r}}
\tableline\tableline
\\ [-2ex]
$f^s$ & $[t^s$&$t^{s+1})$ & &\multicolumn{6}{c}{$-\log_{10}$ (1$-$\SVD{} tolerance)} \\% [1ex]
\cline{5-10}
\\[-2.5ex]
(Hz) & \multicolumn{2}{c}{(s)} & $N^s$ & $1$ & $2$ & $3$ & $4$ & $5$ & $6$ \\ \tableline
4096 & [0&0.5) & 2048 & 1 & 4 & 6 & 8 & 10 & 14 \\
512 & [0.5&4.5) & 2048 & 2 & 6 & 8 & 10 & 12 & 16 \\
256 & [4.5&12.5) & 2048 & 2 & 6 & 8 & 10 & 12 & 15 \\
128 & [12.5&76.5) & 8192 & 6 & 20 & 25 & 28 & 30 & 32 \\
64 & [76.5&140.5) & 4096 & 1 & 8 & 15 & 18 & 20 & 22 \\
64 & [140.5&268.5) & 8192 & 1 & 7 & 21 & 25 & 28 & 30 \\
64 & [268.5&396.5) & 8192 & 1 & 1 & 15 & 20 & 23 & 25 \\
32 & [396.5&460.5) & 2048 & 1 & 1 & 3 & 9 & 12 & 14 \\
32 & [460.5&588.5) & 4096 & 1 & 1 & 7 & 16 & 18 & 21 \\
32 & [588.5&844.5) & 8192 & 1 & 1 & 8 & 26 & 30 & 33 \\
32 & [844.5&1100.5) & 8192 & 1 & 1 & 1 & 12 & 20 & 23 \\
\tableline
\end{tabular}
\end{minipage}
\end{center}
\end{table*}
We use this sub-bank and decomposition for the remainder of this section.

@@

During testing, we were surprised that careful design of the decimation and
interpolation stages made a crucial difference in terms of computational
overhead.  When we tested the pipeline with the largest interpolation filters
provided by GStreamer's \texttt{audioresample} element, we found that rather than
making a negligible contribution to the total \flops\ budget and total measured
overhead they became the dominant contribution.  There is possibly further room
for optimization beyond minimizing $N^\shortuparrow$.  We have considered
designing our own custom decimation and interpolation filters or tuning these
filters separately for each time slice.  We now appreciate that the efficiency
of \lloid\ hinges on proper specification of $N^\shortuparrow$.

\begin{figure*}[b]
	\begin{minipage}[t]{0.5\textwidth}
		\begin{center}
			\includegraphics{f5a}
			(a) Mismatch versus \SVD\ tolerance
		\end{center}
	\end{minipage}
	\begin{minipage}[t]{0.5\textwidth}
		\begin{center}
			\includegraphics{f5b}
			(b) Mismatch versus $N^\shortuparrow$
		\end{center}
	\end{minipage}
	\caption{\label{fig:bw}Box-and-whisker plot of mismatch between nominal
template bank and \lloid\ measured impulse responses.  The upper and lower boundaries of
the boxes show the upper and lower quartiles; the lines in the center denote the medians.
The whiskers represent the minimum and maximum mismatch over all templates.  In 
(a) the interpolation filter length is held fixed at $N^\shortuparrow = 192$, while
the \SVD\ tolerance is varied from $\left(1-10^{-1}\right)$ to $\left(1-10^{-6}\right)$.
In (b), the \SVD\ tolerance is fixed at $\left(1-10^{-6}\right)$ while $N^\shortuparrow$
is varied from 8 to 192.}
\end{figure*}

@@

We are now prepared to offer the estimated computational cost of filtering this
sub-bank of templates compared to other methods.  We used the results of the
previous subsections to set the \SVD\ tolerance to $\left(1-10^{-4}\right)$,
the interpolation filter length to 16, and the decimation filter length to 48.
Table~\ref{table:flops} shows the computational cost in \flops\ for the sub-bank
we described above.  For the \FD\ method, an \fft\ block size of
$\fftblock = 2 \tmpsamps$ is assumed, resulting in a latency of
$\left(\tmpsamps / f^0\right)$ seconds.  Both the \FD\ method and \lloid\ are
five orders of magnitude faster than the conventional \TD\ method.  However, the
\FD\ method has a latency of over half of an hour, whereas the \lloid\ algorithm
has no latency at all.
%
\begin{table*}
\caption{\label{table:flops}Computational cost in \flops\ and latency in seconds
of the \TD\ method, the \FD\ method, and \lloid.  Cost is given
for both the sub-bank described in section~\ref{sec:bank-setup} and a full
\NS--\NS\ search.  The last column gives the approximate number of machines per
detector required for a full advanced LIGO \NS--\NS\ search.}
\begin{center}
\begin{tabular}{lllll}
\tableline\tableline
& \flops\ & & \flops\ & number of \\
method & (sub-bank) & latency (s) & (\NS--\NS) & machines \\[0.1em]
\tableline
time domain & $4.9\times10^{13}$ & 0 & $3.8\times10^{15}$ & $\sim$$10^5$ \\
frequency domain & $5.2\times10^8$ & $2\times10^3$ & $5.9\times10^{10}$ & $\sim$$10$ \\
\lloid\ (theory) & $6.6\times10^8$ & 0 & $\phantom{5.7}\sim 10^{10}$ & $\sim$$10$ \\
\lloid\ (prototype) & (0.9 cores) & $0.5$ & ------------ & $\sim$$10$ \\
\tableline
\end{tabular}
\end{center}
\end{table*}

@@

We wrote a second, stripped pipeline to evaluate the actual latency and
computational overhead.  We executed this pipeline on one of the submit
machines of the \LIGO-Caltech cluster, a Sun Microsystems Sun
Fire\texttrademark\ X4600~M2 server with eight quad-core 2.7~GHz AMD
Opteron\texttrademark\ 8384 processors.  This test consumed $\sim$90\% of the
capacity of just one out of the 32 cores, maintaining a constant latency of
$\sim$0.5~s.

@@

The 0.5~s latency is probably due to buffering and synchronization.  The latency
might be reduced by carefully tuning buffer lengths at every stage in the
pipeline.  Even without further refinements, our implementation of the \lloid\
algorithm has achieved latencies comparable to the \LIGO\ data acquisition
system itself.

\section{Conclusions}

@@

The \lloid{} method is as fast as standard \fft{} convolutions but allows for
latency free, real-time operation.  We anticipate requiring $\sim$40
modern multi-core computers to search for binary neutron stars using
coincident \GW\ data from a 4-detector network.  In the future, additional
computational savings could be acheived by conditionally reconstructing the
\SNR{} time-series only during times when a composite detection statistic
crosses a threshold~\citep{svd-compdetstat}.  However, the anticipated required
number of computers is well within the current computing capabilities of the
\LIGO{} Data Grid.

@@

\LIGO\ was constructed by the California Institute of Technology and
Massachusetts Institute of Technology with funding from the National Science
Foundation and operates under cooperative agreement PHY-0107417.  CH would like
to thank Ilya Mandel for many discussions about rate estimates and the prospects
of early detection, Patrick Brady for countless fruitful conversations about low
latency analysis methods, and John Zweizig for disscussions about \LIGO\ data
acquisition.  NF would like to thank Alessandra Corsi and Larry Price for
illuminating discussions on astronomical motivations.  LS would like thank Shaun
Hooper for productive conversations on signal processing.  This research is
supported by the National Science Foundation through a Graduate Research
Fellowship to LS and by the Perimeter Institute for Theoretical Physics through
a fellowship to CH. DK is supported from the Max Planck Gesellschaft.
